\chapter{Non-deterministic Classical Processes}\label{sec:main}
In this section, we will discuss our main contribution: an extension of \cp
which allows for races while still excluding deadlocks. 
We have seen in \cref{sec:cp-example} how \cp excludes deadlocks, but how exactly
does \cp exclude races?
Let us return to our first example from \cref{sec:introduction}, to the
interaction between John, Mary and the store.
\[
  \begin{array}{c}
    \tm{(\piPar{%
    \piSend{x}{\sliceofcake}{\piSend{x}{\nope}{\store}}
    }{%
    \piPar{\piRecv{x}{y}{\john}}{\piRecv{x}{z}{\mary}}
    })}
    \\[1ex]
    \rotatebox[origin=c]{270}{$\Longrightarrow^{\star}$}
    \\[1ex]
    \tm{(\piPar{\store}{\piPar{\piSub{\sliceofcake}{y}{\john}}{\piSub{\nope}{z}{\mary}}})}
    \quad
    \text{or}
    \quad
    \tm{(\piPar{\store}{\piPar{\piSub{\nope}{y}{\john}}{\piSub{\sliceofcake}{z}{\mary}}})}
  \end{array}
\]
Races occur when more than two processes attempt to communicate simultaneously
over the \emph{same} channel. However, the \text{Cut} rule of \cp requires that
\emph{exactly two} processes communicate over each channel:
\begin{center}
  \cpInfCut
\end{center}
We could attempt write down a protocol for our example, stating that the store
has a pair of channels $\tm{x}, \tm{y} : \ty{\cake}$ with which it communicates
with John and Mary, taking \cake to be the type of interactions in which cake
\emph{may} be obtained, i.e.\ of both \sliceofcake and \nope, and state that the
store communicates with John \emph{and} Mary over a channel of type \ty{\cake
  \parr \cake}.
However, this \emph{only} models interactions such as the following:
\begin{prooftree}
  \AXC{$\seq[{ \john }]{ \Gamma, \tmty{x}{\cake^\bot} }$}
  \AXC{$\seq[{ \mary }]{ \Delta, \tmty{y}{\cake^\bot} }$}
  \SYM{\tens}
  \BIC{$\seq[{ \cpSend{y}{x}{\john}{\mary} }]{
      \Gamma, \Delta, \tmty{y}{\cake^\bot \tens \cake^\bot} }$}
  \AXC{$\seq[{ \store }]{ \Theta, \tmty{x}{\cake}, \tmty{y}{\cake} }$}
  \SYM{\parr}
  \UIC{$\seq[{ \cpRecv{y}{x}{\store} }]{
      \Theta, \tmty{y}{\cake \parr \cake} }$}
  \NOM{Cut}
  \BIC{$\seq[{ \cpCut{y}{\cpSend{y}{x}{\john}{\mary}}{\cpRecv{y}{x}{\store}} }]{
      \Gamma, \Delta, \Theta }$}
\end{prooftree}
Note that in this interaction, John will get whatever the store decides to send
on \tm{x}, and Mary will get whatever the store decides to send on \tm{y}.
This means that this interactions gives the choice of who receives what \emph{to
the store}. This is not an accurate model of our original example, where
the choice of who receives the cake is non-deterministic and depends on factors
outside of any of the participants' control!
And to make matters worse, the term which models our example is entirely
different from the one we initially wrote down in the \textpi-calculus!

The ability to model racy behaviour, such as that in our example, is essential
to describing the interactions that take place in realistic concurrent systems.
Therefore, we would like to extend \cp to allow such races.
Specifically, we would like to do it in a way which mirrors the way in which the
\textpi-calculus handles non-determinism.
We will base our extension on \rcp, a subset of \cp which we introduced in
\cref{sec:background}.
We have chosen to do this to keep our discussion as simple as possible.
Furthermore, as compatibility with the \textpi-calculus is of interest, we will
use the reduction system without commutative conversions, which we introduced in
\cref{sec:cppi}.

This chapter proceeds as follows.
\wen{I should probably write this bit last.}

%% * Terms and types
\section{Terms and types}\label{sec:nc-terms-and-types}
Let us return, briefly, to our example.
\begin{gather*}
  \tm{(\piPar{%
      \piSend{x}{\sliceofcake}{\piSend{x}{\nope}{\store}}
    }{%
      \piPar{\piRecv{x}{y}{\john}}{\piRecv{x}{z}{\mary}}
    })}
\end{gather*}
In this interaction, we see that the channel \tm{x} is used only as a way to
connect the various clients, John and Mary, to the store.
The \emph{real} communication, sending the slice of cake and disappointment,
takes places on the channels \tm{\sliceofcake}, \tm{\nope}, \tm{y} and \tm{z}.

Inspired by this, we add two new constructs to the term language of \cp: sending
and receiving on a \emph{shared} channel.
These actions are marked with a \tm{\star} in order to distinguish them
syntactically from ordinary sending and receiving.
To group clients, we add another form of parallel composition, which we refer to
as \emph{pooling}. 
\input{def-nc-terms}
As before, round brackets denote input, square brackets denote output.
Note that \tm{\ncCnt{x}{y}{P}}, much like \tm{\cpSend{x}{y}{P}{Q}}, is a bound
output---this means that both client creation and server interaction bind a new
name.

In \rcp, we terms are identified up to the commutativity and associativity of
parallel composition. In \nodcap, we add another form of parallel composition,
and therefore must extend our structural congruence:
\input{def-nc-equiv}
We add axioms for the commutativity and associativity of pooling.
We do not add an axiom for \ncEquivPoolAss2, as it follows from
\cref{def:nc-equiv}, see \cref{thm:nc-pool-assoc2}.
It should be noted that \tm{\cpCut{x}{P}{Q}} is considered a single,
\emph{atomic} construct.
Therefore you \emph{cannot} use \ncEquivPoolAss1 to rewrite
\tm{\cpCut{x}{P}{\ncPool{Q}{R}}} to \tm{\cpCut{x}{\ncPool{P}{Q}}{R}}.
We do, however, add two axioms which relate cuts and pool.
We call these \emph{extrusion}, because they closely resemble the
\textpi-calculus axiom for scope extrusion.
We add both \ncRedKappaPool1 and \ncRedKappaPool2, as these relate two different
constructs, and therefore we cannot use the one to derive the other. 
\input{thm-nc-pool-assoc2}
Furthermore, the extensions to structural congruence preserve symmetry.
\input{thm-nc-symmetry}
We can make another observation from our examples.
In every example in which a server interacts with a pool of clients, and which
does not deadlock, there are \emph{exactly} as many clients as there are
server interactions.
Therefore, we add two new \emph{dual} types for client pools and servers, which
track how many clients or server interactions they represent.
\input{def-nc-types}
\input{def-nc-negation}
With these new types, duality remains an involutive function.
\input{thm-nc-negation-involutive}

\section{Typing clients and servers}\label{sec:nc-clients-and-servers}
We have to add typing rules to associate our new client and server
interactions with their types.
The definition for environments will remain unchanged, but we will extend the
definition for the typing judgement.
To determine the new typing rules, we essentially have to answer the question
``What typing constructs do we need to complete the following proof?''
\begin{prooftree}
  \AXC{$\seq[{ \john }]{ \Gamma, \tmty{y}{\cake^\bot} }$}
  \noLine\UIC{$\smash{\vdots}\vphantom{\vdash}$}
  \AXC{$\seq[{ \mary }]{ \Delta, \tmty{y'}{\cake^\bot} }$}
  \noLine\UIC{$\smash{\vdots}\vphantom{\vdash}$}
  \AXC{$\seq[{ \store }]{ \Theta, \tmty{z}{\cake}, \tmty{z'}{\cake} }$}
  \noLine\UIC{$\smash{\vdots}\vphantom{\vdash}$}
  \noLine\TIC{$\seq[{
      \cpCut{x}{\ncPool{\ncCnt{x}{y}{\john}}{\ncCnt{x}{y'}{\mary}}}{
        \ncSrv{x}{z}{\ncSrv{x}{z'}{\store}}} }]{
      \Gamma, \Delta, \Theta }$}
\end{prooftree}
Ideally, we would still like the composition of the client pool and the server
to be a cut. This seems reasonable, as the left-hand side of the term above has
2 clients, and the right-hand side has two server interactions, so \tm{x} is
used at type \ty{\take[2]{\cake^\bot}} on the left, and as \ty{\give[2]{\cake}}
on the right.
\begin{prooftree}
  \AXC{$\seq[{ \john }]{ \Gamma, \tmty{y}{\cake^\bot} }$}
  \noLine\UIC{$\smash{\vdots}\vphantom{\vdash}$}
  \AXC{$\seq[{ \mary }]{ \Delta, \tmty{y'}{\cake^\bot} }$}
  \noLine\UIC{$\smash{\vdots}\vphantom{\vdash}$}
  \noLine\BIC{$\seq[{ \ncPool{\ncCnt{x}{y}{\john}}{\ncCnt{x}{y'}{\mary}} }]{
      \Gamma, \Delta, \tmty{x}{\take[2]{\cake^\bot}} }$}

  \AXC{$\seq[{ \store }]{ \Theta, \tmty{z}{\cake}, \tmty{z'}{\cake} }$}
  \noLine\UIC{$\smash{\vdots}\vphantom{\vdash}$}
  \noLine\UIC{$\seq[{ \ncSrv{x}{z}{\ncSrv{x}{z'}{\store}} }]{
      \Theta, \tmty{x}{\give[2]{\cake}} }$}

  \NOM{Cut}
  \BIC{$\seq[{
      \cpCut{x}{\ncPool{\ncCnt{x}{y}{\john}}{\ncCnt{x}{y'}{\mary}}}{
        \ncSrv{x}{z}{\ncSrv{x}{z'}{\store}}} }]{
      \Gamma, \Delta, \Theta }$}
\end{prooftree}
We will define the typing judgement, and then discuss servers and clients, the
two sides of the above cut, describe the rules we add, and show how they allow
us to complete our proof.
\input{def-nc-typing-judgement}
\input{fig-nc-typing-judgement}

\subsection{Clients and pooling}\label{sec:clients-and-pooling}
A client pool represents a number of independent processes, each wanting to
interact with the server. Examples of such a pool include John and Mary from our
example, customers for online stores in general, and any number of processes
which interact with a single, centralised server.

We introduce two new rules: one to construct clients, and one to pool them
together. The first rule, $(\take[1]{})$, marks interaction over some channel as
a client interaction. It does this by receiving a channel \tm{y} over a
\emph{shared} channel \tm{x}. The channel \tm{y} is the channel across which the
actual interaction will eventually take place.
The second rule, \textsc{Pool}, allows us to pool together clients. This is
implemented, as in the \textpi-calculus, using parallel composition.
\begin{center}
  \ncInfTake1
  \ncInfPool
\end{center}
Using these rules, we can derive the left-hand side of our proof by marking John
and Mary as clients, and pooling them together.
\begin{prooftree}
  \AXC{$\seq[{ \john }]{ \Gamma, \tmty{y}{\cake^\bot} }$}
  \SYM{(\take[1]{})}
  \UIC{$\seq[{ \ncCnt{x}{y}{\john} }]{ \Gamma, \tmty{z}{\take[1]{\cake^\bot}} }$}

  \AXC{$\seq[{ \mary }]{ \Delta, \tmty{y'}{\cake^\bot} }$}
  \SYM{(\take[1]{})}
  \UIC{$\seq[{ \ncCnt{x}{y'}{\mary} }]{ \Delta, \tmty{y'}{\take[1]{\cake^\bot}} }$}

  \NOM{Pool}
  \BIC{$\seq[{ \ncPool{\ncCnt{x}{y}{\john}}{\ncCnt{x}{y'}{\mary}} }]{
      \Gamma, \Delta, \tmty{x}{\take[2]{\cake^\bot}} }$}
\end{prooftree}

\subsection{Servers and contraction}\label{sec:servers-and-contraction}
Dual to a pool of clients is a server. Our interpretation of a server is a
process which offers up some number of interdependent interactions of the same
type. Examples include the store from our example, which gives out slices of
cake and disappointment, online stores in general, and any central server which
interacts with some number of client processes.

We introduce two new rules to construct servers. The first rule, $(\give[1]{})$,
marks a interaction over some channel as a server interaction. It does this by
sending a channel \tm{y} over a \emph{shared} channel \tm{x}. The channel \tm{y}
is the channel across which the actual interaction will eventually take place.
The second rule, \textsc{Cont}, short for contraction, allows us to contract
several server interactions into a single server. This allows us to construct a
server which has multiple interactions of the same type, across the same shared
channel.\footnote{%
  While it ultimately does not matter whether $(\give[1]{})$ and $(\take[1]{})$
  are implemented with a send or a receive action, it feels more natural to have
  the server do the sending.
  Clients indicate their interest in interacting with the server by connecting
  to the shared channel, but it is up to the server to decide \emph{when} to
  interact with each channel.}
\begin{center}
  \ncInfGive1
  \ncInfCont
\end{center}
Using these rules, we can derive the right-hand side of our proof, by marking
each of the store's interactions as server interactions, and then contracting
them.
\begin{prooftree}
  \AXC{$\seq[{ \store }]{ \Theta, \tmty{z}{\cake}, \tmty{z'}{\cake} }$}
  \SYM{(\give[1]{})}
  \UIC{$\seq[{ \ncSrv{x'}{z'}{\store} }]{
      \Theta, \tmty{z}{\cake}, \tmty{x'}{\give[1]{\cake}} }$}
  \SYM{(\give[1]{})}
  \UIC{$\seq[{ \ncSrv{x}{z}{\ncSrv{x'}{z'}{\store}} }]{
      \Theta, \tmty{x}{\give[1]{\cake}}, \tmty{x'}{\give[1]{\cake}} }$}
  \NOM{Cont}
  \UIC{$\seq[{ \ncSrv{x}{z}{\ncSrv{x}{z'}{\store}} }]{
      \Theta, \tmty{x}{\give[2]{\cake}} }$}
\end{prooftree}
Thus, we complete the typing derivation of our example.

\section{Running clients and servers}
Once we have a client/server interaction, how do we run it? Ideally, we would
simply use the reduction rule closest to the one used in the \textpi-calculus. 
\begin{gather*}
  \reducesto
  {\tm{\cpCut{x}{\ncCnt{x}{y}{P}}{\ncSrv{x}{z}{R}}}}
  {\tm{\cpCut{y}{P}{\cpSub{y}{z}{R}}}}
\end{gather*}
However, our case is complicated by the fact that in \tm{\cpCut{x}{P}{Q}} the
name restriction is an inseparable part of the composition, and therefore has to
be part of our reduction rule. 
Because of this, the above reduction can only apply in the singleton case.
If the client pool contains more than one client, such as in the term below,
then there is no way to isolate a single client together with the server,
because \tm{x} occurs in both \tm{\ncCnt{x}{y}{P}} and \tm{\ncCnt{x}{z}{Q}}.
\begin{gather*}
  \tm{\cpCut{x}{\ncPool{\ncCnt{x}{y}{P}}{\ncCnt{x}{z}{Q}}}{\ncSrv{x}{w}{R}}}
  \centernot\Longrightarrow
\end{gather*}
Therefore, we add a second reduction rule, which handles communication between a
one client in a pool of multiple clients and a server.
\begin{gather*}
  \reducesto
  {\tm{\cpCut{x}{\ncPool{\ncCnt{x}{y}{P}}{Q}}{\ncSrv{x}{z}{R}}}}
  {\tm{\cpCut{x}{Q}{\cpCut{y}{P}{\cpSub{y}{z}{R}}}}}
\end{gather*}
Lastly, because we have added another form of parallel composition, we add
another congruence rule, to allow for reduction inside client pools.
\input{def-nc-term-reduction}

The rules \ncRedBetaStar1 and \ncRedBetaStar{n+1} seem like the elimination
rules for a list-like construct. This may come as a surprise, as our client
pools are built up like binary trees, and the typing rules for both sides are
tree-like, with $(\take[1]{})$ and $(\give[1]{})$ playing the role of leaves,
and \textsc{Pool} and \textsc{Cont} merging two trees with $m$ and $n$ leaves
into one with $m+n$ leaves.
However, the server process imposes a sequential ordering on its interactions,
and it is because of this that we have to use list-like elimination rules.

\section{Properties of \nodcap}
In this section, we will revisit the proofs for three important properties of
\rcp, namely preservation, progress, and termination, and show that our
extensions preserve these properties.

\subsection{Preservation}
Preservation is the fact that term reduction preserves typing. There are two
proofs involved in this. First, we show that structural congruence preserves
typing.
\input{thm-nc-preservation-equiv}
\input{fig-nc-preservation-equiv}
Secondly, we prove that term reduction preserves typing.
\input{thm-nc-preservation}
\input{fig-nc-preservation-1}

\subsection{Canonical forms and progress}
In this section, we will extend the definition of canonical forms and the proof
of progress progress given in \cref{sec:cppi}.

\subsubsection{Canonical forms}
First, we extend the definitions of actions with our actions for client and
server creation. 
\input{def-nc-action}
Secondly, as we can reduce inside client pools, we will add pooling to our
definition of evaluation prefixes.
\input{def-nc-evaluation-prefixes}
The definition for the maximum evaluation prefix is unchanged.

There are some subtleties to our definition of canonical forms. The type system
for \cp guarantees that all links directly under an evaluation context act on a
bound channel. Not so for \nodcap.
\begin{scprooftree}
  \AXC{}
  \NOM{Ax}
  \UIC{$\seq[{ \cpLink{x}{y} }]{ \tmty{x}{\take[m]{A}}, \tmty{y}{\give[m]{A^\bot}} }$}
  \AXC{$\seq[{ P }]{ \Gamma, \tmty{x}{\take[n]{A}} }$}
  \NOM{Pool}
  \BIC{$\seq[{ \ncPool{\cpLink{x}{y}}{P} }]{ \Gamma, \tmty{x}{\take[m+n]{A}}, \tmty{y}{\give[m]{A^\bot}} }$}
\end{scprooftree}
There is no way to sensibly reduce this link.
Furthermore, in \cp, if two processes act on the same channel, then they must be
on different sides of the cut introducing that channel. The addition of shared
channels and client pools invalidates this property.
Therefore, we will have to be more careful about the way we define canonical
forms.
We restate the definition of canonical forms below. The additions have been
italicised.
\input{def-nc-canonical-forms}

\subsection{Evaluation contexts}
Evaluation contexts are one-holed term contexts under which reduction can take
place. Since we have added another congruence rule, stating that reduction can
take place inside client pools, we extend our definition of evaluation contexts
to match this.
\input{def-nc-evaluation-contexts}
We also restate \cref{thm:cp-display-cut-1}, and prove that our extension
preserves the property.
\input{thm-nc-display-cut-1}
Furthermore, it will be useful to prove a similar lemma, which shows that we can
push any pooling downwards under an evaluation context.
\input{thm-nc-display-pool-1}

\subsection{Progress}
Progress is the fact that every term is either in some canonical form, or can be
reduced further.
First, we will restate \cref{thm:cp-progress-link} and \cref{thm:cp-progress-beta},
which relate evaluation prefixes and evaluation contexts, and show that our
extension preserves these properties. 
\input{thm-nc-progress-link}
\input{thm-nc-progress-beta}
In essence, \cref{thm:nc-progress-link} and \cref{thm:nc-progress-beta} cover
the cases in which either \cpRedAxCut1 or a \textbeta-reduction rule will be
applied.
However, after applying \cref{thm:nc-progress-beta}, we cannot immediately apply
\ncRedBetaStar{n+1}. For that, we must uncover at least one layer of pooling.
We prove a lemma which states that if we have an interaction on a shared channel
\tm{x}, we can push all pooling rules which pool clients communicating on \tm{x}
inwards. 
\input{thm-nc-progress-shared}
Finally, we are ready to extend our proof of progress. The overall structure of
the proof remains the same, though the addition of pooling makes the wording
slightly more subtle.
\input{thm-nc-progress}

\subsection{Termination}
Termination is the fact that if we iteratively apply progress to obtain a
reduction, and apply that reduction, we will eventually end up with a term in
canonical form.
We restate its proof here for the sake of completeness, but its wording is
unchanged, module references to figures.
\input{thm-nc-termination}

\section{\nodcap and non-deterministic local choice}
In \cref{sec:local-choice}, we discussed the non-deterministic local choice
operator, which is used in several extensions of \piDILL and
\cp~\cite{atkey2016,caires2014,caires2017}.
This operator is admissible in \nodcap.
We can derive the non-deterministic choice \tm{P+Q} by constructing the
following term, provided that \tm{P} and \tm{Q} only communicate over a single
channel: 
\[
  \arraycolsep=0pt
  \tm{
  \begin{array}{lrl}
    \nu x.&((  & \; \ncCnt{x}{y}{\cpCase{y}{P}{\cpHalt{y}}} \\
          &\mid& \; \ncCnt{x}{z}{\cpCase{z}{Q}{\cpHalt{z}}} \; )\\
          &\mid& \; \ncSrv{x}{y}{\ncSrv{x}{z}{\cpInl{y}{\cpWait{y}{\cpInr{z}{\cpLink{z}{w}}}}}} \; )
  \end{array}
  }
\]
The term is a cut between two processes. Let us unpack what each side is doing. 
On the left-hand side, we have a pool with both \tm{P} and \tm{Q} as clients,
wrapped in a case statement which allows the process they are interacting with
to choose whether or not they want them to run. Each of these processes is
constructed as follows: 
\begin{prooftree}
  \AXC{$\seq[{ P }]{ \tmty{y}{A} }$}
  \AXC{}
  \SYM{(\one)}
  \UIC{$\seq[{ \cpHalt{y} }]{ \tmty{y}{\one} }$}
  \SYM{(\with)}
  \BIC{$\seq[{ \cpCase{y}{P}{\cpHalt{y}} }]{ \tmty{y}{A \with \one} }$}
  \SYM{(\take[1]{})}
  \UIC{$\seq[{ \ncCnt{x}{y}{\cpCase{y}{P}{\cpHalt{y}}} }]{
      \tmty{y}{\take[1]{A \with \one}} }$}
\end{prooftree}
On the right-hand side, we have a server which connects to two clients of the
above form, and chooses to terminate one, while allowing the other to run.
This process is constructed as follows:
\begin{prooftree}
  \AXC{}
  \NOM{Ax}
  \UIC{$\seq[{ \cpLink{z}{w} }]{
      \tmty{z}{A^\bot}, \tmty{w}{A} }$}
  \SYM{(\plus_2)}
  \UIC{$\seq[{ \cpInr{z}{\cpLink{z}{w}} }]{
      \tmty{z}{A^\bot \plus \bot}, \tmty{w}{A} }$}
  \SYM{(\bot)}
  \UIC{$\seq[{ \cpWait{y}{\cpInr{z}{\cpLink{z}{w}}} }]{
      \tmty{y}{\bot}, \tmty{z}{A^\bot \plus \bot}, \tmty{w}{A}}$}
  \SYM{(\plus_1)}
  \UIC{$\seq[{ \cpInl{y}{\cpWait{y}{\cpInr{z}{\cpLink{z}{w}}}} }]{
      \tmty{y}{A^\bot \plus \bot}, \tmty{z}{A^\bot \plus \bot}, \tmty{w}{A}}$}
  \SYM{(\give[1]{})}
  \UIC{$\seq[{ \ncSrv{x'}{z}{\cpInl{y}{\cpWait{y}{\cpInr{z}{\cpLink{z}{w}}}}} }]{
      \tmty{y}{(A^\bot \plus \bot)}, \tmty{x'}{\give[1]{(A^\bot \plus \bot)}}, \tmty{w}{A} }$}
  \SYM{(\give[1]{})}
  \UIC{$\seq[{ \ncSrv{x}{y}{\ncSrv{x'}{z}{\cpInl{y}{\cpWait{y}{\cpInr{z}{\cpLink{z}{w}}}}}} }]{
      \tmty{x}{\give[1]{(A^\bot \plus \bot)}}, \tmty{x'}{\give[1]{(A^\bot \plus \bot)}}, \tmty{w}{A} }$}
  \NOM{Cont}
  \UIC{$\seq[{ \ncSrv{x}{y}{\ncSrv{x}{z}{\cpInl{y}{\cpWait{y}{\cpInr{z}{\cpLink{z}{w}}}}}} }]{
      \tmty{x}{\give[2]{(A^\bot \plus \bot)}, \tmty{w}{A}}}$}
\end{prooftree}
When we compose these two processes, we get a process of the type
$\seq[{ P + Q}]{ \tmty{w}{A} }$, using \tm{P+Q} as an abbreviation for the
encoding of non-deterministic local choice.

The full version of local choice, which allows you to choose between two
processes with an arbitrary number of channels, is admissible.
We can construct it using the invertibility of $(\parr)$, the logical property
that we can turn any context $\tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}$ into a single
formula $\tmty{x}{A_1 \parr \dots \parr A_n}$, and vice versa.
In terms of the process calculus, this means that for any number of channels
there exist processes which can systematically pack or unpack those channels.
\begin{lemmaB}
  If $\seq[{ P }]{ \Gamma, \tmty{x_1}{A_1}\dots\tmty{x_n}{A_n} }$, then
  $\seq[{ \cpRecv{x_1}{x_n}{\dots\cpRecv{x_1}{x_2}{P}} }]{
    \Gamma, \tmty{x_1}{A_1 \parr \dots \parr A_n} }$.
\end{lemmaB}
\begin{proof}
  By repeated application of $(\parr)$.
\end{proof}

\begin{lemmaB}
  If $\seq[{ \cpRecv{x_1}{x_n}{\dots\cpRecv{x_1}{x_2}{P}} }]{
    \Gamma, \tmty{x_1}{A_1 \parr \dots \parr A_n} }$, then there exists a term
  \tm{P'} such that
  $\seq[{ P' }]{ \Gamma, \tmty{x_1}{A_1}\dots\tmty{x_n}{A_n}}$
  and $\tm{P'} \Longrightarrow^{3(n-1)} \tm{P}$.
\end{lemmaB}
\begin{proof}
  By induction on $n$.
  \begin{itemize}
  \item
    Case $n = 1$.
    We have $\seq[{ P }]{ \Gamma, \tmty{x_1}{A_1} }$.
    This is our desired result.
  \item
    Case $n = k + 1$. We have:
    \[
      \seq[{
      \cpRecv{x_1}{x_n}{\cpRecv{x_1}{x_k}{\dots\cpRecv{x_1}{x_2}{P}}} }]{
      \Gamma, \tmty{x_1}{A_1 \parr \dots \parr A_k \parr A_n} }.
    \]
    We construct the process:

    \[
      \tm{\cpCut{x_1}{
          \cpRecv{x_1}{x_n}{\cpRecv{x_1}{x_k}{\dots\cpRecv{x_1}{x_2}{P}}}}{
          \cpSend{x_1}{x_n}{\cpLink{x_n}{x_n'}}{\cpLink{x_1}{x_1'}}
        }}.
    \]
    In three steps, this reduces to
    \(
      \tm{\cpRecv{x_1'}{x_k}{\dots\cpRecv{x_1'}{x_2}{\cpSub{x_1'}{x_1}{\cpSub{x_n'}{x_n}{P}}}}}.
    \)
    We then substitute \tm{x_1} for \tm{x_1'} and \tm{x_n} for \tm{x_n'}, and
    apply the induction hypothesis.
  \end{itemize}
\end{proof}
While there is a certain amount of overhead involved in this encoding, it scales
linearly in terms of the number of processes.
The reverse---encoding the non-determinism present in \nodcap using
non-deterministic local choice---scales exponentially, as with the \textpi-calculus.


\wen{Add a discussion of non-determinism in \nodcap versus non-determinism in
  the \textpi-calculus. Mention at least that terms which have clients who are
  and who aren't blocked on an external communication will prefer those clients
  who aren't blocked when non-deterministically reducing.}
%%% Local Variables:
%%% TeX-master: "main"
%%% End:
