\chapter{\cp as a type system for the \textpi-calculus}\label{sec:cppi}
\cp has a tight correspondence with classical linear logic.
This has many advantages.
It is deadlock free and terminating, and as seen in \cref{sec:cp-properties},
the proofs of its meta-theoretical properties are brief and aesthetically
pleasing.

The price paid for this is a somewhat weaker correspondence with the
\textpi-calculus.
It would be useful to be able to think of \cp as a type system for the
\textpi-calculus.
However, as it stands there are many differences between these systems.
Most prominent of these are the commutative conversions. These reduction rules
are taken directly from the proof normalisation procedure of classical linear
logic, and do not correspond to any reductions in the \textpi-calculus.

\citenat{lindley2015semantics} observed that, using a different reduction
strategy, which more closely resembles that of the \textpi-calculus, we can
ensure that the commutative conversions are always applied \emph{last}.
That is to say, they define two separate reduction relations:
$\longrightarrow_{C}$ for \cpRedAxCut1, \cpRedAxCut2 and \textbeta-reductions,
and $\longrightarrow_{CC}$ for commutative conversions, and show that any
sequence of reductions has the following form:
\[
  P \longrightarrow_{C} \dots \longrightarrow_{C} Q \longrightarrow_{CC} \dots \longrightarrow_{CC} R
\]
In this dissertation, we use a reduction strategy which follows
\citenat{lindley2015semantics}, but opts to drop the suffix of commutative
conversions.
As a consequence of this, we can drop the commutative conversions from our
reduction system, which therefore more tightly corresponds to that of the
\textpi-calculus. 
A consequence of dropping the commutative conversions is that the reduction
system as a whole becomes deterministic.
The price we pay for this is a weaker correspondence to classical linear logic.
This shows in our notion of canonical form, which is weaker, and in our proof of
progress, which is much more involved.

For clarity's sake: whenever we refer to \cp or \rcp, for the remainder of this
dissertation, we refer to the variant \emph{without} commutative conversions,
i.e.\ which uses the following definition of term reduction.
\input{def-cp-term-reduction-2}

This chapter will proceed as follows.
In \cref{sec:cp-canonical-forms}, we will describe what it means for a term to
be in canonical form.
In \cref{sec:cp-evaluation-contexts}, we will define evaluation contexts.
In \cref{sec:cp-progress}, we will give a new proof of progress, which follows
\citenat{lindley2015semantics}.

\section{Canonical forms}\label{sec:cp-canonical-forms}
The reduction strategy described by \citeauthor{lindley2015semantics} applies
\cpRedAxCut1, \cpRedAxCut2, and \textbeta-reductions until the process blocks on
one or more external communications, and then applies the commuting conversions
to bubble one of those external communications to the front of the term.
This allows them to define canonical forms as any term which is not a cut.
For us, terms in canonical form will be those terms which are blocked on an
external communication, before any commutative conversions are applied.
In this section, we will describe the form of such terms.

We have informally used the phrase ``act on'' in previous sections. It is time
to formally define what it means when we say a process \emph{acts on} some
channel.
\input{def-cp-action}
Furthermore, we will need the notion of an \emph{evaluation prefix}.
Intuitively, evaluation prefixes are multi-holed contexts consisting solely of
cuts. We will use evaluation prefixes in order to have a view of every
\emph{action} in a process at once.
\input{def-cp-evaluation-prefixes}
Intuitively, we can say that every term of the form
\tm{\cpPlug{G}{P_1 \dots P_n}} is equivalent to some term of the form
\tm{\cpCut{x_1}{P_1}{\cpCut{x_2}{P_2}{\dots \cpCut{x_n}{P_{n-1}}{P_n} \dots}}} 
where $\tm{x_1} \dots \tm{x_{n-1}}$ are the channels bound in \tm{G}.
In fact, a similar equivalence was used by \citeauthor{lindley2015semantics}
\cite{lindley2015semantics} in their semantics for \cp. 
\input{def-cp-maximum-evaluation-prefix}
We can now define what it means for a term to be in canonical form. Intuitively,
a process is in canonical form either when there is no top-level cut, or when it
is blocked on an external communication. We state this formally as follows:
\input{def-cp-canonical-forms}
This definition is adequate, as it matches our intuition. We will see this in
our proof for \cref{thm:cp-progress-3}.

\section{Evaluation contexts}\label{sec:cp-evaluation-contexts}
Intuitively, evaluation contexts are one-holed term contexts under which
reduction can take place. For \rcp, these consist solely of cuts.
\input{def-cp-evaluation-contexts}
We can prove that we can push any cut downwards under an evaluation context, as
long as the channel it binds does not occur in the context itself.
\input{thm-cp-display-cut-1}
And vice versa. However, we will not use the following lemma in this
dissertation, and leave its proof as an exercise to the reader.
\input{thm-cp-display-cut-2}

\section{Progress}\label{sec:cp-progress}
Progress is the fact that every term is either in some canonical form, or can be
reduced further.
%
There are two important lemmas which we will need in order to prove progress.
These relate evaluation prefixes to evaluation contexts.
%
Specifically, if a process under an evaluation prefix is a link, we can rewrite
the entire process in such a way as to reveal the cut which introduced one of
the channels acted upon by that link.
\input{thm-cp-progress-link}
And if two processes under an evaluation prefix act on the same channel, then we
can rewrite the entire process in such a way as to reveal the cut which
introduced that channel. 
\input{thm-cp-progress-beta}

Finally, we are ready to prove progress.
Note that in proofs throughout this dissertation, we will leave uses of
\cpRedGammaEquiv and the congruence rules for reductions implicit. 
\input{thm-cp-progress-3}
The proof of progress described in this section is novel, though it takes
inspiration from the reduction system for \cp described by \citenat{lindley2015semantics}. 
The proof of itself is somewhat involved. The reason for
this is that we want our reduction strategy to match that of the
\textpi-calculus as closely as possible.
It is also for this reason that our proof of progress involves some
non-determinism. For instance, in the second case of the proof, we do not
specify how to select the two processes \tm{P_i} and \tm{P_j} if there are
multiple options available.

\section{Example}
Let's have a look at the differences between the reduction strategy we have
defined in this chapter, and the reduction strategy which \citenat{wadler2012}
defines. Let's imagine the following scenario:
\begin{quote}
  Alice, John, and Mary went on a lovely trip together.
  However, John and Mary can be a bit scatterbrained sometimes, and it just so
  happens that they both forgot to bring their wallets.
  They both owe Alice some money, which they're now trying to pay back at the
  same time.
\end{quote}
We can model this interaction in \cp as the following term, assuming \alice,
\john, and \mary are three processes representing Alice, John and Mary, and
\bankjohn and \bankmary are two processes representing John and Mary's
respective banks.
Note that, in an abuse of notation, we will adopt an ``anti-Barendregt''
convention for our examples.
This means that if two channel names will be identified after a reduction, we
will preemptively give them the same name, for instance, \tm{z} and \tm{w} in
the example below. 
\begin{align*}
  \tm{\cpCut{x}{\cpCut{y}{
  \cpRecv{x}{z}{\cpRecv{y}{w}{\alice}}}{
  \cpSend{y}{w}{\bankjohn}{\john}}}{
  \cpSend{x}{z}{\bankmary}{\mary}}}
\end{align*}
Note that in the above interaction, no \textbeta-reduction rule applies
\emph{immediately}. This means that using the reduction strategy described by
\citenat{wadler2012}, we will apply a commutative conversion. 
\begin{align*}
  \begin{array}{lll}
    \tm{\cpCut{x}{\cpCut{y}{
    \cpRecv{x}{z}{\cpRecv{y}{w}{\alice}}}{
    \cpSend{y}{w}{\bankjohn}{\john}}}{
    \cpSend{x}{z}{\bankmary}{\mary}}}
    & \Longrightarrow & \text{by \cpRedKappaParr}
    \\
    \tm{\cpCut{x}{\cpRecv{x}{z}{\cpCut{y}{
    \cpRecv{y}{w}{\alice}}{
    \cpSend{y}{w}{\bankjohn}{\john}}}}{
    \cpSend{x}{z}{\bankmary}{\mary}}}
    & \Longrightarrow & \text{by \cpRedBetaTensParr}
    \\
    \tm{\cpCut{z}{\cpCut{x}{\cpCut{y}{
    \cpRecv{y}{w}{\alice}}{
    \cpSend{y}{w}{\bankjohn}{\john}}}{
    \mary}}{
    \bankmary}}
    & \Longrightarrow & \text{by \cpRedBetaTensParr}
    \\
    \tm{\cpCut{z}{\cpCut{x}{\cpCut{w}{\cpCut{y}{
    \alice}{
    \john}}{
    \bankjohn}}{
    \mary}}{
    \bankmary}}
  \end{array}
\end{align*}
On the other hand, using the reduction strategy described in this chapter, we will
rewrite using structural congruence.
\begin{align*}
  \begin{array}{lll}
    \tm{\cpCut{x}{\cpCut{y}{
    \cpRecv{x}{z}{\cpRecv{y}{w}{\alice}}}{
    \cpSend{y}{w}{\bankjohn}{\john}}}{
    \cpSend{x}{z}{\bankmary}{\mary}}}
    & \equiv          & \text{by \cref{thm:cp-display-cut-1}}
    \\
    \tm{\cpCut{y}{\cpCut{x}{
    \cpRecv{x}{z}{\cpRecv{y}{w}{\alice}}}{
    \cpSend{x}{z}{\bankmary}{\mary}}}{
    \cpSend{y}{w}{\bankjohn}{\john}}}
    & \Longrightarrow & \text{by \cpRedBetaTensParr}
    \\
    \tm{\cpCut{y}{\cpCut{z}{\cpCut{x}{
    \cpRecv{y}{w}{\alice}}{
    \mary}}{
    \bankmary}}{
    \cpSend{y}{w}{\bankjohn}{\john}}} 
    & \equiv          & \text{by \cref{thm:cp-display-cut-1}}
    \\
    \tm{\cpCut{z}{\cpCut{x}{\cpCut{y}{
    \cpRecv{y}{w}{\alice}}{
    \cpSend{y}{w}{\bankjohn}{\john}}}{
    \mary}}{
    \bankmary}}
    & \Longrightarrow & \text{by \cpRedBetaTensParr}
    \\
    \tm{\cpCut{z}{\cpCut{x}{\cpCut{w}{\cpCut{y}{
    \alice}{
    \john}}{
    \bankjohn}}{
    \mary}}{
    \bankmary}}
  \end{array}
\end{align*}
Note that the second reduction sequence is a \emph{valid} reduction sequence in
the either reduction system---it is simply not the sequence \emph{chosen} by the
reduction strategy described by \citenat{wadler2012}.

%%% Local Variables:
%%% TeX-master: "main"
%%% End:
