\documentclass[a4paper,UKenglish]{lipics-v2016}
\bibliographystyle{plainurl}%
% Author macros::begin
\title{Give or Take: Races in Classical Linear Logic}%
\titlerunning{Give or Take}%
\author[1]{Pepijn Kokke}%
\author[1]{J. Garrett Morris}%
\author[1]{Philip Wadler}%
\affil[1]{LFCS, University of Edinburgh, Edinburgh, United Kingdom\\
  \texttt{\{pepijn.kokke, garrett.morris, philip.wadler\}@ed.ac.uk}}%
\authorrunning{P. Kokke, J. G. Morris, and P. Wadler}%
\Copyright{Pepijn Kokke, J. Garrett Morris, and Philip Wadler}%
\subjclass{
  F.4.1. [Mathematical Logic]: Lambda calculus and related systems;
  D.3.1. [Formal Definitions and Theory]: Syntax;
  D.3.2. [Language Classifications] Concurrent, distributed, and parallel languages
}%
\keywords{$\pi$-calculus, linear logic, session types, deadlock freedom, non-determinism}
% Author macros::end

% Editor-only macros::begin
\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important (Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
% Editor-only macros::end

% Preamble macros::begin
\input{preamble}
% Preamble macros::end

\begin{document}

\maketitle

\begin{abstract}
  Process calculi based in logic, such as CP, provide a foundation for
  deadlock-free concurrent programming, but at the cost of excluding
  non-determinism and races.  We introduce \nodcap (nodcap), which extends CP
  with a novel account of non-determinism.  Our approach draws on bounded linear
  logic to provide a strongly-typed account of standard process calculus
  expressions of non-determinism.  We show that our extension is expressive
  enough to capture many uses of non-determinism in untyped calculi, such as
  non-deterministic choice, while preserving CP's metatheoretic properties,
  including deadlock freedom.  We have formalized our calculus and its
  properties using Agda.
\end{abstract}

\section{Introduction}\label{sec:introduction}
Imagine the following scenario:
\begin{quote}
  John and Mary are working from home one morning when they each get a craving
  for a slice of cake. Being denizens of the web, they quickly find a nearby
  p\^atisserie that does home deliveries. Unfortunately for them, they both
  order their cake at the \emph{same} p\^atisserie, which has only one slice
  left. After that, all it can deliver is disappointment.
\end{quote}
This scenario is a classic example of a race condition, and it is modeled in the
$\pi$-calculus by the program below, given that \john, \mary, and \ptis\ are
processes representing John, Mary and the p\^atisserie, respectively, and
\cake\ and \nocake\ are two channels over which \ptis\ will send a slice of
cake, and disappointment, respectively.
The arrows indicate two possible reductions of the program: one in which John
gets the cake and Mary is disappointed, and one in which Mary gets cake and John
is disappointed.
\def\exampleprograma{\red{\ensuremath{%
      \cut{x}(\recv{x}{y}\john \mid \recv{x}{z}\mary \mid
      \send{x}{\cake}\send{x}{\nocake}\ptis)
    }}}
\def\exampleresultaa{\red{\ensuremath{%
      \cut{x}(\subst{\john}{\cake}{y} \mid \subst{\mary}{\nocake}{z} \mid \ptis)
    }}}
\def\exampleresultab{\red{\ensuremath{%
      \cut{x}(\subst{\john}{\nocake}{y} \mid \subst{\mary}{\cake}{z} \mid \ptis)
    }}}
\def\examplereductiona{%
  \begin{tikzpicture}
    \node (program) at (0,0) {$\exampleprograma\quad$};
    \node[above=1cm, right=1cm, at=(program.east)] (resulta)
    {$\!\!^\star\quad\exampleresultaa$};
    \node[below=1cm, right=1cm, at=(program.east)] (resultb)
    {$\!\!^\star\quad\exampleresultab$};
    \draw[->] (program.east) to [out=045,in=180] (resulta.west);
    \draw[->] (program.east) to [out=315,in=180] (resultb.west);
  \end{tikzpicture}
}
\[\examplereductiona\]
Races show up in any client/server system where clients compete for limited
server resources. It is therefore important that any concurrent programming
language allows us to write them.
\\
As shown by the above example, the $\pi$-calculus is up to the
task of modeling such programs. However, it can also model some undesirable
programs. For example, consider the
following scenario:
\begin{quote}
  John managed to order the last slice of cake from the p\^atisserie. However,
  Mary \emph{really} wants it, and offers him some money for it. John agrees,
  but sadly there is no trust between the two. Mary demands to get the cake
  first, before she gives John the money. John demands to get the money first,
  before he gives Mary the cake.
\end{quote}
This is a classic example of a deadlock, and it is modeled in the
$\pi$-calculus by the program below, given that \john\ and \mary\ are processes
representing John and Mary, respectively, \cake\ is a channel over which
\john\ will send a slice of cake, and \dollar\ is a channel over which \mary\ %
will send money.
\def\exampleprogramb{\red{\ensuremath{%
      \cut{x}(\recv{x}{y}\send{x}{\cake}\john \mid \recv{x}{z}\send{x}{\dollar}\mary)
    }}}
\[\exampleprogramb\]
The program is stuck---John is refusing to send the cake until he receives money
from Mary, and Mary is refusing to send the money until John sends the cake, so
nothing will actually happen.
It is undesirable in the sense that deadlock is unlikely to be the intended
behaviour of a program.

Session types~\cite{honda1993} provide static guarantees that concurrent
processes respect specified communication protocols.
Recently, work on
$\pi\text{DILL}~$\cite{caires2010} and CP~\cite{wadler2012} has yielded tight
correspondences between session typed $\pi$-calculi and intuitionistic and
classical linear logic~\cite{girard1987}, respectively.
These not only
interpret session types as linear propositions, but also show a
correspondence between reduction in the $\pi$-calculus and cut elimination in
linear logic.
Furthermore, by the very nature of their correspondence with linear logic,
programs in these calculi are guaranteed to be free from deadlocks.
Unfortunately, they are also deterministic---i.e., unable to model races.

We present \nodcap (nodcap), an extension of CP~\cite{wadler2012} with
a novel account of non-determinism and races.  Inspired by bounded linear
logic~\cite{girard1992}, we introduce a form of shared channels, in which the
type of a shared channel tracks how many times it is reused.  As in the untyped
$\pi$-calculus, sharing introduces the potential of non-determinism.  We show
that our approach is sufficient to capture practical examples of races, such as
the p\^atisserie, as well as other formal characterizations of non-determinism,
such as non-deterministic choice.  However, \nodcap does not abandon the
metatheoretical benefits of CP: we show that it enjoys cut elimination, and
so \nodcap processes are still guaranteed to be deadlock-free.

The paper proceeds as follows.  Section~\ref{sec:background} introduces our
variant of the $\pi$-calculus and its type system CP~\cite{wadler2012}.
Section~\ref{sec:races} introduces \nodcap, our extension to CP; shows that cut
elimination for CP extends to \nodcap, guaranteeing deadlock freedom; and
discusses the types of concurrency captured by \nodcap.
Section~\ref{sec:conclusion} concludes with discussions of related and future work.
%
Our type system and cut elimination procedure
have been formalised using Agda~\cite{norell2009}, and the code is
available on GitHub.\footnote{%
  \url{https://github.com/pepijnkokke/nodcap}
}

\section{Background}
\label{sec:background}

In this section, we will discuss a rudimentary subset of the session-typed
process calculus CP~\cite{wadler2012}, which we will refer to as RCP.
Under the correspondence with linear logic, RCP corresponds to
multiplicative-additive linear logic.
The process calculus used by RCP is a variant of the
$\pi$-calculus~\cite{milner1992b}, defined by
the following grammar:
\[
  \begin{aligned}
    \red{P},\:&\red{Q},\red{R} ::=\\
    &       \red{\link{x}{y}}           &&\text{link}
    &\qquad&\red{\cut{x}(P \mid Q)}     &&\text{parallel composition}\\
    &       \red{\send{x}{y}(P \mid Q)} &&\text{``output''}
    &\qquad&\red{\recv{x}{y}P}          &&\text{``input''}\\
    &       \red{\inl{x}P}              &&\text{left selection}
    &\qquad&\red{\inr{x}P}              &&\text{right selection}\\
    &       \red{\case{x}{P}{Q}}        &&\text{choice}
    &\qquad&\red{\case{x}{}{}}          &&\text{empty choice}\\
    &       \red{\send{x}{}0}           &&\text{empty output}
    &\qquad&\red{\recv{x}{}P}           &&\text{empty input}\\
  \end{aligned}
\]
The construct $\link{x}{y}$ links two channels~\cite{sangiorgi1996,boreale1998},
forwarding messages received on $x$ to $y$ and vice versa.
%
The construct $\cut{x}{(P \mid Q)}$ creates a new channel $x$, and composes two
processes, which communicate on $x$, in parallel.
Therefore, in $\cut{x}{(P \mid Q)}$ the name $x$ is bound in both $P$ and $Q$.
%
In $\recv{x}{y}P$ and $\send{x}{y}P$, round brackets are used for input, square
brackets for output.
We use bound output~\cite{sangiorgi1996}---this means that unlike in the
$\pi$-calculus, both input and output bind a new name.
In $\recv{x}{y}P$ the new name $y$ is bound in $P$.
In $\send{x}{y}{(P \mid Q)}$, the new name $y$ is only bound in $P$, while $x$
is only bound in $Q$.

Processes in RCP are typed using session types which correspond to the
multiplicative and additive connectives from linear logic. These are defined
using the following grammar:
\[
  \begin{aligned}
    \blue{A},\:&\blue{B},\blue{C} ::=\\
    &       \tm{A \tens B} &&\text{`tensor', independent sessions A and B}
    &\qquad&\tm{\one}      &&\text{unit for \tens}\\
    &       \tm{A \parr B} &&\text{`par', possibly dependent sessions A and B}
    &\qquad&\tm{\bot}      &&\text{unit for \parr}\\
    &       \tm{A \plus B} &&\text{`plus', select session A or B}
    &\qquad&\tm{\nil}      &&\text{unit for \plus}\\
    &       \tm{A \with B} &&\text{`with', offer session A or B}
    &\qquad&\tm{\top}      &&\text{unit for \with}
  \end{aligned}
\]
We discuss the interpretations of the connectives below, together with the
typing rules which introduce them.

Environments associate channels with types. They are defined as follows:
\[
  \blue{\Gamma},\blue{\Delta},\blue{\Theta} ::=
  \tm[x_1]{A_1},\tm[x_2]{A_2},\dots,\tm[x_n]{A_n}
\]
Names in an environment must be unique.
and environments $\Gamma$ and $\Delta$ can
only be combined ($\Gamma, \Delta$) if $\text{fv}(\Gamma) \cap
\text{fv}(\Delta) = \varnothing$.

Judgements $\seq[{ P }]{ \tm[x_1]{A_1},\tm[x_2]{A_2},\dots,\tm[x_n]{A_n} }$
denote that process $P$ communicates along channels $x_i$ following the
communication protocol specified by $A_i$.
The multiplicatives ($\tens$, $\parr$) deal with dependence and independence.
A channel of type $A \tens B$ represents a pair of channels which communicate
with two \emph{independent} processes---that is to say, two processes who share
no channels.
A channel of type $A \parr B$ represents a pair of channels which are used
within a single process. This means that their interactions can depend on one
another---i.e.\ the interaction of type $B$ could depend on the result of the
interaction of type $A$, or vice versa.
While the rules for $\tens$ and $\parr$ introduce input and output operations,
these are inessential---the essential distinction lies in the two independent
processes in $\tens$ as opposed to the single process in $\parr$.
\begin{center}
  \begin{prooftree*}
    \AXC{$\seq[ P ]{ \Gamma , \tm[y]{A} }$}
    \AXC{$\seq[ Q ]{ \Delta , \tm[x]{B} }$}
    \SYM{\tens}
    \BIC{$\seq[{ \send{x}{y}(P \mid Q) }]{ \Gamma , \Delta , \tm[x]{A \tens B} }$}
  \end{prooftree*}
  \begin{prooftree*}
    \AXC{$\seq[ P ]{ \Gamma , \tm[y]{A} , \tm[x]{B} }$}
    \SYM{\parr}
    \UIC{$\seq[ \recv{x}{y}P ]{ \Gamma , \tm[x]{A \parr B} }$}
  \end{prooftree*}
\end{center}
The multiplicative units ($\one$, $\bot$) follow the same pattern.
A process with a channel of type $\one$ has no independent processes, and so
must halt.
A process with a channel of type $\bot$ is not further restricted.
\begin{center}
  \begin{prooftree*}
    \AXC{}
    \SYM{\one}
    \UIC{$\seq[{ x[].0 }]{ \tm[x]{\one} }$}
  \end{prooftree*}
  \begin{prooftree*}
    \AXC{$\seq[P]{ \Gamma }$}
    \SYM{\bot}
    \UIC{$\seq[{x().P}]{ \Gamma , \tm[x]{\bot} }$}
  \end{prooftree*}
\end{center}
The additives ($\plus$, $\with$) deal with choice.
A process communicating over a channel of type $A \plus B$ can select either an
interaction of type $A$ or one of type $B$.
A process communicating over a channel of type $A \with B$ has to
offer an interaction either of type $A$ or of type $B$.
\begin{center}
  \begin{prooftree*}
    \AXC{$\seq[ P ]{ \Gamma , \tm[x]{A} }$}
    \SYM{\plus_1}
    \UIC{$\seq[{ \inl{x}{P} }]{ \Gamma , \tm[x]{A \plus B} }$}
  \end{prooftree*}
  \begin{prooftree*}
    \AXC{$\seq[ P ]{ \Gamma , \tm[x]{A} }$}
    \AXC{$\seq[ Q ]{ \Delta , \tm[x]{B} }$}
    \SYM{\with}
    \BIC{$\seq[ \case{x}{P}{Q} ]{ \Gamma , \Delta , \tm[x]{A \with B} }$}
  \end{prooftree*}
\end{center}
The additive units ($\nil$, $\top$) follow the same pattern, except with nullary
instead of binary choices. Along a channel of type $\nil$, a
process has to select one of \emph{zero} options.
As this is impossible, there is no rule for $\nil$.
When communicating over a channel of type $\top$, a process has to be prepared
to offer any of \emph{zero} different interactions, and so is
trivial.
\begin{center}
  \begin{prooftree*}
    \AXC{}
    \SYM{\top}
    \UIC{$\seq[ \case{x}{}{} ]{ \tm[x]{\top} }$}
  \end{prooftree*}
\end{center}
Duality plays a crucial role in linear logic and session types. It ensures that
a process offering a choice on one end of a channel is matched with a process
which accepts a choice on the other. It is also crucial to deadlock freedom, as
it ensures that a process which uses communication on $x$ to decide what to send
on $y$ is communicating with two independent processes on $x$ and $y$,
preventing circular dependencies. Each type $A$ has a dual $A^\bot$, defined as follows:
\[
  \begin{array}{ccccccc}
    \tm{( A \tens B )^\bot} = \tm{A^\bot \parr B^\bot}&&
    \tm{\one^\bot}          = \tm{\bot}&&
    \tm{( A \parr B )^\bot} = \tm{A^\bot \parr B^\bot}&&
    \tm{\bot^\bot}          = \tm{\one}\\
    \tm{( A \plus B )^\bot} = \tm{A^\bot \with B^\bot}&&
    \tm{\nil^\bot}          = \tm{\top}&&
    \tm{( A \with B )^\bot} = \tm{A^\bot \plus B^\bot}&&
    \tm{\top^\bot}          = \tm{\nil}
  \end{array}
\]
Duality appears in the typing rules for two CP forms.  Forwarding $\link x y$
connects two channels with dual endpoints, while composition $\cut x (P \mid Q)$
requires that $P$ and $Q$ use $x$ with dual expectations.
\begin{center}
  \begin{prooftree*}
    \AXC{$\vphantom{yA^\bot}$}
    \NOM{Ax}
    \UIC{$\seq[{ \link{x}{y} }]{ \tm[x]{A}, \tm[y]{A^\bot} }$}
  \end{prooftree*}
  \begin{prooftree*}
    \AXC{$\seq[{ P }]{ \Gamma, \tm[x]{A} }$}
    \AXC{$\seq[{ Q }]{ \Delta, \tm[y]{A^\bot} }$}
    \NOM{Cut}
    \BIC{$\seq[{ \cut{x}(P \mid Q) }]{ \Gamma, \Delta }$}
  \end{prooftree*}
\end{center}

\noindent
We return to our example of a deadlock:
\[\exampleprogramb\]
As CP and RCP are deadlock free, there should be no valid type
assignment for this program.
There are, however, two close variants which are typable.
These correspond to the programs in which John decides
which of the goods is sent first, and that in which Mary makes the decision.
The first of these is modeled in RCP by the program below, given that $\john$ is
a process which will receive money and send out cake, in any order, $\mary$ is a
process which will receive cake, $\dollar$ is a process which will send money,
$\money$ the type of money, $\plato$ is the type of $\cake$, and the
environments $\Gamma$, $\Delta$ and $\Theta$ summarise the remaining
interactions of John, Mary and the p\^atisserie with the rest of the world.
\begin{prooftree}
  \AXC{$\seq[{\john}]{\Gamma, \tm[y]{\money^\bot}, \tm[x]{\plato}}$}
  \SYM{\parr}
  \UIC{$\seq[{\recv{x}{y}\john}]{\Gamma, \tm[x]{\money^\bot \parr \plato}}$}
  \AXC{$\seq[{\dollar}]{\Delta, \tm[z]{\money}}$}
  \AXC{$\seq[{\mary}]{\Theta, \tm[x]{\plato^\bot}}$}
  \SYM{\tens}
  \BIC{$\seq[{\send{x}{z}(\dollar \mid \mary)}]{\Delta, \tm[x]{\money \tens \plato^\bot}}$}
  \NOM{Cut}
  \BIC{$\seq[{\cut{x}(\recv{x}{y}\john \mid \send{x}{z}(\dollar \mid \mary))}]{\Gamma, \Delta, \Theta}$}
\end{prooftree}
Note that the typing rule for $\tens$ forces Mary, who in our example sends
money to John and receives cake in return, to be represented by two
\emph{independent} processes $\dollar$ and $\mary$, one of which sends money, and
the other receives cake.\footnote{%
  There is a third variant of the deadlocked process which is harmless but
  \emph{not} well-typed in RCP. This is the variant in which neither John
  nor Mary decides which of the goods is sent first. This would
  be equivalent to assigning the types $\money^\bot \tens \plato$ and $\money
  \tens \plato^\bot$ to John and Mary, respectively. Such a program becomes
  typable in a version of CP extended with \textsc{Mix}~\cite{atkey2016}.
}

Under the correspondence with linear logic, communication in RCP is identified
with the elimination of the \textsc{Cut} rule. We can show that any term can be
systematically rewritten into a normal form in which all cuts are eliminated. We
present a proof of this here for reference, as we will extend it to cover
\nodcap in later sections:
\begin{quote}
  \textit{Proof sketch.}
  Given a program with one or more cuts, we select any such application $\cut{x}(P
  \mid Q)$ such that $P$ and $Q$ are in normal form.
  There are three cases:
  \begin{enumerate}
  \item
    If either $P$ or $Q$ is a link $\link{x}{y}$ we can eliminate the cut
    immediately, e.g.\ %
    \begin{center}
      \begin{prooftree*}
        \AXC{}
        \NOM{Ax}
        \UIC{$\seq[{ \link{y}{x} }]{ \tm[y]{A^\bot}, \tm[x]{A} }$}
        \AXC{$\seq[{ P }]{ \Gamma, \tm[x]{A^\bot} }$}
        \NOM{Cut}
        \BIC{$\seq[{ \cut{x}(\link{y}{x} \mid P) }]{ \Gamma, \tm[y]{A^\bot} }$}
      \end{prooftree*}
      $\Longrightarrow$
      \begin{prooftree*}
        \AXC{$\seq[{ \subst{P}{y}{x} }]{ \Gamma, \tm[y]{A^\bot} }$}
      \end{prooftree*}
    \end{center}
  \item
    If the last rules applied in $P$ and $Q$ both introduced the top-level
    connective in the type of $x$, then we apply any of the principal cut
    reductions, seen in Figure~\ref{fig:cp-cut-prin}. These replace a cut with
    one or more cuts on channels with a \emph{strictly smaller} type.
  \item
    Otherwise, we apply one of the commutative conversions for \textsc{Cut},
    seen in Figure~\ref{fig:cp-cut-cc}.\footnote{%
      In the interest of space, we are presenting the commutative conversions on
      terms alone.
    }
    These move the application of \textsc{Cut} further into the term.
  \end{enumerate}
  We iteratively apply this procedure until there are no more applications of
  \textsc{Cut} left in the program.
\end{quote}
The proof above differs from the proof given by Wadler~\cite{wadler2012}, which
only reduces top-level cuts in order to be consistent with operational
accounts of the $\pi$-calculus.
%
We present the full version here as we intend to show that \nodcap is consistent
as a logic as well as deadlock-free as a programming language.
%
We recover a version consistent with operational accounts of $\pi$-calculus by
introducing structural congruences reflecting the associativity and
commutativity of \textsc{Cut}:
\begin{align*}
  \red{\cut x (P \mid Q)} &\equiv \red{\cut x (Q \mid P)} \\
  \red{\cut x (P \mid \cut y (Q \mid R))} &\equiv \red{\cut y (Q \mid \cut x (P \mid R))} && \text{if $x \not\in \text{fv}(Q)$} \\
  \red{\cut x (P \mid R)} &\equiv \red{\cut x (Q \mid R)} &&\text{if $P \equiv Q$}
\end{align*}
and by only selecting cuts nested under other applications of \textsc{Cut}.
%
\input{fig-cp-cut-prin}
\input{fig-cp-cut-cc}


\section{Races in CP}
\label{sec:races}

We return to our initial example to see how RCP excludes races:
\[\examplereductiona\]
Races occur when more than two processes attempt to communicate simultaneously
over the same channel.
However, the \textsc{Cut} rule requires that \emph{exactly two} processes
communicate over each channel:
% GM: not sure that repeating the rule here is necessary.
\begin{prooftree}
  \AXC{$\seq[{P}]{\Gamma,\tm[x]{A}}$}
  \AXC{$\seq[{Q}]{\Delta,\tm[x]{A^\bot}}$}
  \NOM{Cut}
  \BIC{$\seq[{\cut{x}(P \mid Q)}]{\Gamma,\Delta}$}
\end{prooftree}
Modelling racy behaviour, such as that in our example, is essential to
describing realistic concurrent systems. Therefore, we would like to
extend RCP to accept such races. This essentially boils down to the
following question---what typing constructs do we need to complete the following
proof:\footnote{%
  It should be noted that $\cake$ and $\nocake$ are both considered to be of the
  type \plato, with $\nocake$ being the vastly more disappointing value of the
  cake type.
}
\begin{prooftree}
  \AXC{$\seq[{ \john }]{ \Gamma, \tm[y]{\plato^\bot} }$}
  \AXC{$\seq[{ \mary }]{ \Delta, \tm[z]{\plato^\bot} }$}
  \AXC{$\seq[{ \ptis }]{ \Theta, \tm[\cake]{\plato}, \tm[\nocake]{\plato} }$}
  \noLine\TIC{$\vdots$}\noLine
  \UIC{$\seq[{ \exampleprograma }]{ \Gamma, \Delta, \Theta }$}
\end{prooftree}
Ideally, we would use a regular \textsc{Cut} to introduce the channel $x$ in
this proof. We can do so by grouping John and Mary on one side of the
\textsc{Cut}, and putting the p\^atisserie on the other side. This seems
reasonable---John and Mary both try to get cake, and the p\^atisserie offers
to try to provide cake twice.
\begin{prooftree}
  \AXC{$\seq[{ \john }]{ \Gamma, \tm[y]{\plato^\bot} }$}
  \noLine\UIC{$\vdots$}\noLine
  \AXC{$\seq[{ \mary }]{ \Delta, \tm[z]{\plato^\bot} }$}
  \noLine\UIC{$\vdots$}\noLine
  \BIC{$\seq[{ (\recv{x}{y}\john \mid \recv{x}{z}\mary) }]{ \Gamma, \Delta, \tm[x]{???^\bot} }$}
  \AXC{$\seq[{ \ptis }]{ \Theta, \tm[\cake]{\plato}, \tm[\nocake]{\plato} }$}
  \noLine\UIC{$\vdots$}\noLine
  \UIC{$\seq[{ \send{x}{\cake}\send{x}{\nocake}\ptis }]{\Theta, \tm[x]{???} }$}
  \NOM{Cut}
  \BIC{$\seq[{ \exampleprograma }]{ \Gamma, \Delta, \Theta }$}
\end{prooftree}
Furthermore, the actions of John and Mary are completely independent, whereas
for the p\^atisserie its actions are order dependent---it will first ship a slice of
cake, and from then on then ship only disappointment.
This makes it seem like we could fill in the $???$ with the type
$\plato^\bot \parr \plato^\bot$.
\def\exampleprogramc{\red{\ensuremath{%
      \cut{x}(\send{x}{y}(\john \mid \mary) \mid \recv{x}{z}\ptis)
    }}}%
We would implement our ``race'' with the program below:
\begin{prooftree}
  \AXC{$\seq[{ \john }]{ \Gamma, \tm[y]{\plato^\bot} }$}
  \AXC{$\seq[{ \mary }]{ \Delta, \tm[x]{\plato^\bot} }$}
  \SYM{\tens}
  \BIC{$\seq[{ \send{x}{y}(\john \mid \mary) }]{ \Gamma, \Delta,
      \tm[x]{\plato^\bot \tens \plato^\bot} }$}
  \AXC{$\seq[{ \ptis }]{ \Theta, \tm[x]{\plato}, \tm[z]{\plato} }$}
  \SYM{\parr}
  \UIC{$\seq[{ \recv{z}{x}\ptis }]{ \Theta, \tm[x]{\plato \parr \plato} }$}
  \NOM{Cut}
  \BIC{$\seq[{ \exampleprogramc }]{ \Gamma, \Delta, \Theta }$}
\end{prooftree}
However, this program gives the choice of who receives the cake to the
p\^atisserie. This no longer accurately models our original example, where the
choice of who receives the cake is non-deterministic, and depends on factors
outside of the p\^atisserie's control (i.e.\ who happens to place their
order first).

Thus, we add a new dual pair of connectives to RCP---$\give[n]{A}$ (give) and
$\take[n]{A}$ (take). Like an $n$-ary par, $\give[n] A$ denotes a channel with
$n$ dependent $A$ channels, and $\take[n] A$ denotes a channel with $n$
independent $A$ channels, like an $n$-ary tensor.  However, unlike the case for
the $\parr/\tens$ reduction, the reduction of $\give[n] A$ and $\take[n] A$ is
non-deterministic.  The indexing by positive natural numbers $n$ assures that
there are sufficient server resources for each client, and that server resources
are used completely.

In the following two subsections, we will look at the two sides of the
\textsc{Cut} in our incomplete proof, describe the rules we add for `give' and
`take', and show how they allow us to complete our proof.  Then, we show that
our proof of cut elimination of CP extends to \nodcap, and discuss the forms of
concurrency expressible with \nodcap.

\subsection*{Servers and contraction}
Our interpretation of a server is any process which offers up a sequence of
identically typed interactions which are dependent on one another. Examples include
our p\^atisserie, which gives out cake until it runs out or a ticket server which
hands out numbered tickets in order.
We introduce two new rules to construct servers. The first, $\give[1]{}$, marks
a channel as a server interaction. It does this by sending the channel $y$
across a channel $x$.\footnote{%
  Ultimately, it does not matter whether this rule is implemented with a send or
  a receive action. All that matters is that the operations introduced by
  $\give[1]{}$ and $\take[1]{}$ are dual.
}
The second rule, \textsc{Cont}, enables the contraction of server interactions.
This enables us to construct a server which has multiple interactions of the
same type across the same channel $x$.
\begin{center}
  \begin{prooftree*}
    \AXC{$\seq[{P}]{ \Gamma , \tm[y]{A} }$}
    \SYM{\give[1]{}}
    \UIC{$\seq[{ \give{\send{x}{y}}P }]{ \Gamma , \tm[x]{\give[1]{A}} }$}
  \end{prooftree*}
  \begin{prooftree*}
    \AXC{$\seq[{ P }]{ \Gamma , \tm[x]{\give[m]{A}} , \tm[y]{\give[n]{A}} }$}
    \NOM{Cont}
    \UIC{$\seq[{ \subst{P}{x}{y} }]{ \Gamma , \tm[x]{\give[m+n]{A}} }$}
  \end{prooftree*}
\end{center}
Using these rules, we can derive the right-hand side of our proof by marking
each of the p\^atisserie's interactions as server interactions, and then
contracting them:
\begin{prooftree}
  \AXC{$\seq[{ \ptis }]{ \Theta, \tm[\cake]{\plato}, \tm[\nocake]{\plato} }$}
  \SYM{\give[1]{}}
  \UIC{$\seq[{ \give{\send{x'}{\nocake}}\ptis }]{
      \Theta, \tm[\cake]{\plato}, \tm[x']{\give[1]{\plato}} }$}
  \SYM{\give[1]{}}
  \UIC{$\seq[{ \give{\send{x}{\cake}}\give{\send{x'}{\nocake}}\ptis }]{
      \Theta, \tm[x]{\give[1]{\plato}}, \tm[x']{\give[1]{\plato}} }$}
  \NOM{Cont}
  \UIC{$\seq[{ \give{\send{x}{\cake}}\give{\send{x}{\nocake}}\ptis }]{
      \Theta, \tm[x]{\give[2]{\plato}} }$}
\end{prooftree}


\subsection*{Clients and pooling}
The dual of a server is not a client, but a pool of clients---a number of
independent processes which each wish to interact with the
server. Examples include our processes John and Mary or customers to a ticket server.
We introduce two new rules---one to construct clients and one to
\emph{pool} them together. The first, $\take[1]{}$, marks an interaction as a
client interaction. It does this by receiving the channel $y$, across which the
actual interaction will take place, over some shared channel $x$.
The second rule, \textsc{Pool}, enables us to pool together clients. This is
implemented using a $\pi$-calculus parallel composition, which puts together two
client pools, in which each client attempts to interact across $x$.
\begin{center}
  \begin{prooftree*}
    \AXC{$\seq[{ P }]{ \Gamma , \tm[y]{A} }$}
    \SYM{\take[1]{}}
    \UIC{$\seq[{ \take{\recv{x}{y}}P }]{ \Gamma , \tm[x]{\take[1]{A}} }$}
  \end{prooftree*}
  \begin{prooftree*}
    \AXC{$\seq[{ P }]{ \Gamma , \tm[x]{\take[m]{A}} }$}
    \AXC{$\seq[{ Q }]{ \Delta , \tm[x]{\take[n]{A}} }$}
    \NOM{Pool}
    \BIC{$\seq[{ (P \mid Q) }]{ \Gamma , \Delta , \tm[x]{\take[m+n]{A}} }$}
  \end{prooftree*}
\end{center}
Using these rules, we can derive the left-hand side of our proof by marking John
and Mary as clients looking for cake, and pooling them together:
\begin{prooftree}
  \AXC{$\seq[{ \john }]{ \Gamma, \tm[y]{\plato^\bot} }$}
  \SYM{\take[1]{}}
  \UIC{$\seq[{ \take{\recv{x}{y}}\john }]{ \Gamma, \tm[x]{\take[1]{\plato^\bot}} }$}
  \AXC{$\seq[{ \mary }]{ \Delta, \tm[z]{\plato^\bot} }$}
  \SYM{\take[1]{}}
  \UIC{$\seq[{ \take{\recv{x}{z}}\mary }]{ \Delta, \tm[x]{\take[1]{\plato^\bot}} }$}
  \SYM{}
  \BIC{$\seq[{ (\take{\recv{x}{y}}\john \mid \take{\recv{x}{z}}\mary) }]{
      \Gamma, \Delta, \tm[x]{\take[2]{\plato^\bot}} }$}
\end{prooftree}
Thus, we complete the derivation of our example.

\subsection*{Non-deterministic communication and cut elimination}
All that remains is to show that the proof of cut elimination for RCP can be
extended to cover \nodcap. In order to show this, we need to do two things.
We need to
\begin{enumerate}
\item
  give \emph{one} new principal cut reduction, to show that we can reduce a cut
  on a channel of type $\give[n]{A}$/$\take[n]{A}$ to a number of smaller cuts.
\item
  give \emph{four} new commutative conversion, in order to show that
  \textsc{Cut} commutes with the four new proof rules we have introduced; and
\end{enumerate}
The commutative conversions are shown in Figure~\ref{fig:nodcap-cut}.
%
\input{fig-nodcap-cut}
%
All that remains is to give a reduction rule for a cut between a server and a
client pool:
\begin{prooftree}
  \AXC{$\seq[{ P }]{ \Gamma, \tm[x]{\give[n]{A^\bot}}}$}
  \AXC{$\seq[{ R }]{ \Delta, \tm[x]{\take[n]{A}} }$}
  \NOM{Cut}
  \BIC{$\seq[{ \cut{x}(P \mid R) }]{ \Gamma, \Delta }$}
\end{prooftree}
In order to do this, we introduce two new admissible rules---server expansion,
and client interleaving:
\begin{center}
  \begin{scprooftree*}[0.90]
    \AXC{$\seq[{ P }]{ \Gamma, \tm[x]{\give[n]{A}} }$}
    \NOM{Exp}
    \UIC{$\seq[{ \expn{x}{x_1 \cdots x_n}P }]{ \Gamma, \tm[x_1]{A} \cdots \tm[x_n]{A} }$}
  \end{scprooftree*}%
  \begin{scprooftree*}[0.90]
    \AXC{$\seq[{ P }]{ \Delta, \tm[y]{\take[n]{A}} }$}
    \AXC{$\seq[{ Q }]{ \Gamma, \tm[x_1]{A^\bot} \cdots \tm[x_n]{A^\bot} }$}
    \NOM{Int}
    \BIC{$\seq[{ \intl{y}{x_1 \cdots x_n}(P \mid Q) }]{ \Gamma, \Delta }$}
  \end{scprooftree*}
\end{center}
We can compose these rules to eliminate the above cut (see
Figure~\ref{fig:nodcap-cut}).
%
However, this has simply moved the problem: How do we eliminate applications of
\textsc{Exp} and \textsc{Int}? This involves two more proofs with the exact same
structure as that of cut elimination.

The principal eliminations and commutative conversions for \textsc{Exp} are
shown in Figure~\ref{fig:nodcap-expn}. When we put these together, we see that
the operation $\expn{x}{x_1 \dots x_n}P$ strips all the applications of
$\give[1]{}$ and \textsc{Cont} from a program, and renames the channels which
were marked as server interactions using the names $x_1 \dots x_n$.
%
\input{fig-nodcap-expn}
%
The principal eliminations and commutative conversions for \textsc{Int} are
shown in Figure~\ref{fig:nodcap-intl}. When we put these together, we see that
the operation $\intl{x}{y_1 \dots y_n}(P \mid Q)$ joins each interaction $y_i$
with some dual interaction from the client pool. This means that each
interleaving reduces to $n$ cuts, where $n$ is the number of clients.
%
\input{fig-nodcap-intl}

So, how does the non-determinism arise? All the interactions in our cut are
of the same type. When eliminating the cut, there is an intrinsic
choice: it does not matter which client we connect to which server
interaction. Therefore, we apply a permutation $\pi$ to the names in our cut
reduction (see Figure~\ref{fig:nodcap-cut}).

Let us go back to our example. We can now assign a type to this program, and
consider its reduction behaviour as a result of cut elimination.
Figure~\ref{fig:cut-ex} shows both possible reductions of this term.
%
\input{fig-cut-ex}

As for RCP, we can also describe a version of cut reduction that does not reduce
under prefixes, matching operational accounts of $\pi$-calculus.  To do so,
we would introduce structure congruences for \textsc{Cut} as before, and
restrict the algorithm to only select cuts nested under applications of
\textsc{Cont}, \textsc{Pool}, or other applications of \textsc{Cut}.


\subsection*{\nodcap and non-deterministic choice}
Recent work~\cite{caires2014,caires2017,atkey2016} has seen the extension of
$\pi\text{DILL}$ and CP with non-deterministic
choice~\cite{milner1992b}.
This operator can essentially be summarised by the following typing rule and
reduction semantics:
\begin{center}
  \begin{prooftree*}
    \AXC{$\seq[{P}]{\Gamma}$}
    \AXC{$\seq[{Q}]{\Gamma}$}
    \BIC{$\seq[{P+Q}]{\Gamma}$}
  \end{prooftree*}
  \hspace*{2cm}
  \(\!
  \begin{aligned}
    \red{P+Q} \Longrightarrow \red{P}\\
    \red{P+Q} \Longrightarrow \red{Q}
  \end{aligned}
  \)
\end{center}
This operator is derivable in \nodcap. We can derive the non-deterministic
choice $P+Q$ for a single channel by combining the following two terms with
\textsc{Cut}:
\begin{prooftree}
  \AXC{$\seq[{P}]{\Gamma, \tm[y]{A}}$}
  \SYM{\with}
  \UIC{$\seq[{\case{y}{\halt{y}}{P}}]{
      \Gamma, \tm[y]{\one \with A}}$}
  \SYM{\take[1]{}}
  \UIC{$\seq[{\take{\recv{x}{y}}\case{y}{\halt{y}}{P}}]{
      \Gamma, \tm[x]{\take[1]{(\one \with A)}}}$}
  \AXC{$\seq[{Q}]{
      \Delta, \tm[z]{A}}$}
  \SYM{\with}
  \UIC{$\seq[{\case{z}{\halt{z}}{Q}}]{
      \Delta, \tm[z]{\one \with A}}$}
  \SYM{\take[1]{}}
  \UIC{$\seq[{\take{\recv{x}{z}}\case{z}{\halt{z}}{Q}}]{
      \Delta, \tm[x]{\take[1]{(\one \with A)}}}$}
  \NOM{Pool}
  \BIC{$\seq[{(
      \take{\recv{x}{y}}\case{y}{\halt{y}}{P} \mid
      \take{\recv{x}{z}}\case{z}{\halt{z}}{Q}
      )}]{\Gamma, \Delta, \tm[x]{\take[2]{\one \with A}}}$}
\end{prooftree}
\begin{prooftree}
  \AXC{}
  \NOM{Ax}
  \UIC{$\seq[{
      \link{z}{w}
    }]{\tm[z]{A^\bot}, \tm[w]{A}}$}
  \SYM{\plus_2}
  \UIC{$\seq[{
      \inr{z}\link{z}{w}
    }]{\tm[z]{\bot \plus A^\bot}, \tm[w]{A}}$}
  \SYM{\bot}
  \UIC{$\seq[{
      \wait{y}\inr{z}\link{z}{w}
    }]{\tm[y]{\bot}, \tm[z]{\bot \plus A^\bot}, \tm[w]{A}}$}
  \SYM{\plus_1}
  \UIC{$\seq[{
      \inl{y}\wait{y}\inr{z}\link{z}{w}
    }]{\tm[y]{\bot \plus A^\bot}, \tm[z]{\bot \plus A^\bot}, \tm[w]{A}}$}
  \SYM{\give[1]{(2)}}
  \UIC{$\seq[{
      \give{\send{x}{y}}\give{\send{x'}{z}}\inl{y}\wait{y}\inr{z}\link{z}{w}
    }]{\tm[x]{\give[1]{(\bot \plus A^\bot)}}, \tm[x']{\give[1]{(\bot \plus A^\bot)}}, \tm[w]{A}}$}
  \NOM{Cont}
  \UIC{$\seq[{
      \give{\send{x}{y}}\give{\send{x}{z}}\inl{y}\wait{y}\inr{z}\link{z}{w}
    }]{\tm[x]{\give[2]{(\bot \plus A^\bot)}, \tm[w]{A}}}$}
\end{prooftree}
And we can derive full non-deterministic choice using the invertability of
$\parr$.


\section{Related and Future Work}
\label{sec:conclusion}

\subsection*{Relation to bounded linear logic}
We mentioned in Section~\ref{sec:introduction} that \nodcap was inspired by
bounded linear logic~(BLL)~\cite{girard1992}. BLL is a typed lambda calculus
based on intuitionistic linear logic which guarantees that its programs are
polynomial-time functions.
It too uses resource-indexed exponentials. However, instead of interpreting
these as client and server interactions, BLL interprets them as accesses to a
memory cell, as is a common interpretation in linear logic~\cite{girard1987}.
There are some superficial differences between BLL and \nodcap, e.g.\ the former
is intuitionistic while the latter is classical, but the main difference between
the two lies in storage versus pooling. In BLL, $\take[n]{A}$ denotes a memory
cell which can be accessed $n$ times, whereas in \nodcap, $\take[n]{A}$
represents a pool of $n$ different values, computed independently by $n$
different processes.

\subsection*{Recursion and resource variables}

Our formalism so far has only captured servers that provide for a fixed number
of clients.  More realistically, we would want to define servers that provide
for arbitrary numbers of clients.  This poses two problems: how would we define
arbitrarily-interacting stateful processes, and how would we extend the
typing discipline of \nodcap to account for them without losing its static
guarantees.

One approach to defining server processes would be to combine \nodcap with
structural recursion and corecursion, following the $\mu\text{CP}$ extension of Lindley
and Morris~\cite{lindley2016}.  Their approach can express processes which
produce streams of $A$ channels; such a process would expose a channel with the
corecursive type $\nu X. A \parr (1 \plus X)$.  Given such a process, it is
possible to produce a channel of type $A \parr A \parr \cdots \parr A$ for any
number of $A$s, allowing us to satisfy the type $\give[n]{A}$ for an arbitrary
$n$.

We would also need to extend the typing discipline to capture arbitrary use of
shared channels.  One approach would be to introduce resource variables and
quantification.  Following this approach, in addition to having types $\give[n]
A$ and $\take[n] A$ for concrete $n$, we would also have types $\give[x] A$ and
$\take[x] A$ for resource variables $x$.  These variables would be introduced by
quantifiers $\forall x A$ and $\exists x A$.  Defining terms
corresponding to $\forall x A$, and its relationship with structured recursion,
presents an interesting area of further work.


\subsection*{Relation to exponentials in CP}

Our account of CP has not included the exponentials $\give A$ and $\take A$.
The type $\take A$ denotes arbitrarily many independent instances of $A$, while
the type $\give A$ denotes a concrete (if unspecified) number of
potentially-dependent instances of $A$.  Existing interpretations of linear
logic as session types have taken $\take A$ to denote $A$-servers, while
$\give A$ denotes $A$-clients.  However, the analogy is imperfect: while we
expect servers to provide arbitrarily many instances of their behavior, we also
expect those instances to be interdependent.

With quantification over resource variables, we can give precise accounts
of both CP's exponentials and idealised servers and clients.  CP
exponentials could be embedded into this framework using the definitions
$\take{A} := \forall{n}\take[n]{A}$ and $\give{A} := \exists{n}{\give[n]{A}}$.
We would also have types that precisely matched our intuitions for server and
client behavior: an $A$ server is of type $\forall n \give[n] A$, being
unbounded but dependent, while a collection of $A$ clients is of type
$\exists n \take[n] A$, being definitely sized by independent.

\clearpage
\bibliography{main}

\end{document}
%%% Local Variables:
%%% TeX-master: "main"
%%% fill-column: 80
%%% End:
